---
title: "Comp/Stat 112: Bivariate Visualizations"
author: "INSERT STUDENT NAME HERE (COLLABORATOR NAMES HERE)"
output:
  bookdown::html_document2:
    split_by: none
    toc: yes
    toc_depth: 3
    toc_float:
      toc_collapsed: true
    number_sections: false
---

```{r 04_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(ggplot2)
library(fivethirtyeight)
```

# Bivariate Visualizations


## Getting to know the dataset {-}

```{exercise}
Begin by loading the data from "https://www.macalester.edu/~ajohns24/data/electionDemographics16.csv" and getting to know the data. Follow the prompts below to guide you.
```

> ANSWER: See code below.

```{r eval=TRUE}
# Load data from "https://www.macalester.edu/~ajohns24/data/electionDemographics16.csv"
elect <- read_csv("https://www.macalester.edu/~ajohns24/data/electionDemographics16.csv")


# Check out the first rows of elect.  What are the units of observation?


# How much data do we have?


# What are the names of the variables?
```



```{exercise}
Explore the win column:
    The `winrep_2016` variable indicates whether or not the Republican (Trump) won the county in 2016, thus is *categorical*.  Let's construct both numerical and visual summaries of Trump wins/losses.  (Before you do, what do you anticipate?) 
```

> ANSWER: 

```{r eval=TRUE}
# Construct a table (a numerical summary) of the number of counties that Trump won/lost
table(xxx) # fill in the xxx

# Attach a library needed for ggplots
library(xxx)

# Construct a bar chart (a visual summary) of this variable.
ggplot(xxx, aes(xxx)) +
  geom_xxx()
```

```{exercise name="Explore Vote Percentages"}
The `perrep_2016` variable includes a bit more detail about Trump's support in each county.   

```

a. Since it's *quantitative* we need different tools to visually explore the variability in `perrep_2016`.  To this end, construct & interpret both a histogram and density plot of `perrep_2016`.  (Before you do, what do you anticipate?)  

> ANSWER (interpretation): 

```{r eval=TRUE}
# histogram
ggplot(elect, aes(xxx)) +
  geom_xxx(color = "white")

# density plot
ggplot(elect, aes(xxx)) +
  geom_xxx()
```   

b. Thus far, we have a good sense for how Trump’s support varied from county to county. We don’t yet have a good sense for *why*. What other variables (ie. county features) might explain some of the variability in Trump’s support from county to county?  Which of these variables do you think will be the best predictors of support?  The worst?    

> ANSWER (list ideas): 


## Background on visualizing relationships {-}


```{exercise, name="Mock-Ups"}
Consider a subset  of the variables: 

```

```{r echo=FALSE, warning=FALSE}
fd <- elect %>%
  filter(region %in% c(8039, 28003, 40129, 29119, 13247, 27027)) %>%
  select(c(county, abb, perrep_2016, perrep_2012, winrep_2016, StateColor)) %>%
  mutate(perrep_2016 = round(perrep_2016, 2), perrep_2012 = round(perrep_2012, 2))
head(fd)
```

In groups, sketch on paper a mock-up of a visualization of the relationship between the given pair of variables (i.e., what type of chart is appropriate to demonstrate the relationship?):

a. The relationship between `perrep_2016` (the response) and `perrep_2012` (the predictor).  

> ANSWER [describe sketch]: 

b. The relationship between `perrep_2016` (the response) and `StateColor` (the predictor).  Think: how might we modify the below density plot of `perrep_2016` to distinguish between counties in red/purple/blue states?  

> ANSWER [describe sketch]: 

    
```{r fig.width=2, fig.height=2}
ggplot(elect, aes(x = perrep_2016)) +
  geom_density()
```    

c. The relationship between Trump's county-levels wins/losses `winrep_2016` (the response) and `StateColor` (the predictor).  Think: how might we modify the below bar plot of `winrep_2016` to distinguish between counties in red/purple/blue states? 

> ANSWER [describe sketch]: 

```{r fig.width=3, fig.height=2}
ggplot(elect, aes(x = winrep_2016)) +
  geom_bar()
```    

## Visualizing quantitiative vs quantitative relationships {-}

Let's start by exploring the relationship between Trump's 2016 support (`perrep_2016`) and Romney's 2012 support (`perrep_2012`), both quantitative variables.    

```{exercise name="Scatterplots and Glyphs"}
Both `perrep_2016` and `perrep_2012` are quantitative, thus require their own axes.  Traditionally, the response variable (what we are trying to predict or explain) is placed on the y-axis.  Once the axes are set up, each case is represented by a "glyph" at the coordinates defined by these axes.    

```

a. Make a scatterplot of `perrep_2016` vs `perrep_2012` with different glyphs: points or text.    
  
> ANSWER:  Run code below and look.

```{r eval=TRUE}
# just a graphics frame
ggplot(elect, aes(y = perrep_2016, x = perrep_2012))

# add a layer with "point" glyphs
ggplot(elect, aes(y = perrep_2016, x = perrep_2012)) +
  geom_point()

# add a layer with symbol glyphs
ggplot(elect, aes(y = perrep_2016, x = perrep_2012)) +
  geom_point(shape = 3)

# add a layer with "text" glyphs
ggplot(elect, aes(y = perrep_2016, x = perrep_2012)) +
  geom_text(aes(label = abb))
```    

b. Summarize the relationship between the Republican candidates' support in 2016 and 2012.  Be sure to comment on:    
        - the strength of the relationship (weak/moderate/strong)    
        - the direction of the relationship (positive/negative)    
        - outliers (In what state do counties deviate from the national trend?  Explain why this might be the case)    

> ANSWER (summarize): 

```{exercise name="Capture the Trend with 'smooths'"}
The trend of the relationship between `perrep_2016` and `perrep_2012` is clearly positive and (mostly) linear.  We can highlight this trend by adding a model "smooth" to the plot.    
```

a. Add a layer with a model smooth:  
 
> ANSWER: Run Code Below and look.

```{r eval=TRUE}
ggplot(elect, aes(y = perrep_2016, x = perrep_2012)) +
  geom_point() +
  geom_smooth()
```

b. Construct a new plot that contains the model smooth but does not include the individual cases (eg: point glyphs).    

> ANSWER: Write Code Below and note the impact on the graphic.

```{r}

```

c. Notice that there are gray bands surrounding the blue model smooth line.  What do these gray bars illustrate/capture and why are they widest at the "ends" of the model?  

> ANSWER: 


By default, `geom_smooth` adds a smooth, localized model line.  To examine the "best" *linear model*, we can specify `method="lm"`:    
  
```{r eval=TRUE}
ggplot(elect, aes(y = perrep_2016, x = perrep_2012)) +
  geom_point() +
  geom_smooth(method = "lm")
```    


```{exercise name="Modify the Scatterplots"}
As with univariate plots, we can change the aesthetics of scatterplots.    
```

a. Add the following:
  - Add appropriate axis labels to your scatterplot.  Label the y-axis "Trump 2016 support (%)" and label the x-axis "Romney 2012 support (%)".    
  - Change the color of the points.    
  - Add some *transparency* to the points.  NOTE: `alpha` can be between 0 (complete transparency) and 1 (no transparency).      
b. Why is transparency useful in this particular graphic?
   
   
> ANSWER (b in sentence form):    

```{r, eval=TRUE}
#a)


```

    
```{exercise name="More Scatterplots"}
2012 results aren't the only possible predictor of 2016 results.  Consider two more possibilities.    
```

a. Write code to construct a scatterplot of `perrep_2016` and `median_rent`.  Summarize the relationship between these two variables.    

> ANSWER (summarize):  


```{r, eval=TRUE}
#a)


```

b. Write code to construct a scatterplot of `perrep_2016` and `percent_white`.  Summarize the relationship between these two variables.    

> ANSWER (summarize):


```{r, eval=TRUE}
#b)

```


c. Among `perrep_2012`, `median_rent` and `percent_white`, which is the best predictor of `perrep_2016`?  Why? 

> ANSWER:  




### Visualizing quantitative vs. categorical relationships {-}

To visualize the relationship between Trump's 2016 support (`perrep_2016`) and the `StateColor` (categorical) we need to incorporate a grouping mechanism.  Work through the several options below.    

```{exercise name="Side-by-Side Density Plots"}
We can show density plots for each state color next to each other:
```

a. Construct a density plot for each group.  

> ANSWER: Run Code Below.

```{r eval=TRUE}
ggplot(elect, aes(x = perrep_2016, fill = StateColor)) +
  geom_density()
```    
    
b.  Notice that `ggplot` randomly assigns colors to group based on alphabetical order.  In this example, the random color doesn't match the group itself (red/purple/blue)!  We can fix this:  

> ANSWER:  Run Code Below.

```{r eval=TRUE}
ggplot(elect, aes(x = perrep_2016, fill = StateColor)) +
  geom_density() +
  scale_fill_manual(values = c("blue", "purple", "red"))
```    
    
c. The overlap between the groups makes it difficult to explore the features of each.  One option is to add *transparency* to the density plots:    

> ANSWER: Run Code Below. 

```{r eval=TRUE}
ggplot(elect, aes(x = perrep_2016, fill = StateColor)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "purple", "red"))
```    

d. Yet another option is to separate the density plots into separate "facets" defined by group:    

> ANSWER:  Run Code Below.

```{r eval=TRUE}
ggplot(elect, aes(x = perrep_2016, fill = StateColor)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "purple", "red")) +
  facet_wrap(~ StateColor)
```
    

```{exercise name="Side-by-Side Histograms"}
Let's try a similar strategy using histograms to illustrate the relationship between `perrep_2016` and `StateColor`.    
```

a. Start with the default histogram:    

> ANSWER:  Run Code Below.

```{r eval=TRUE}
ggplot(elect, aes(x = perrep_2016, fill = StateColor)) +
  geom_histogram(color = "white")
```

b. That's not very helpful!  Separate the histograms into separate facets for each `StateColor` group.    

> ANSWER: Add Code Below. 

```{r eval=TRUE}
#b)

```

```{exercise name="More Options!"}
Density plots and histograms aren't the only type of viz we might use...    
```

a. Construct side-by-side violins and side-by-side boxplots (see description below).  

> ANSWER:  Run Code Below.

```{r eval=TRUE}
# violins instead
ggplot(elect, aes(y = perrep_2016, x = StateColor)) +
  geom_violin()

# boxes instead
ggplot(elect, aes(y = perrep_2016, x = StateColor)) +
  geom_boxplot()
```    

Box plots are constructed from five numbers - the minimum, 25th percentile, median, 75th percentile, and maximum value of a quantitative variable. 
    
b. In the future, we'll typically use *density plots* instead of histograms, violins, and boxplots.  Explain at least one pro and one con of the density plot.    
  
> ANSWER:


```{exercise}
Let's not forget the most important purpose of these visualizations!  Summarize the relationship between Trump's 2016 county-level support among red/purple/blue states.  
```

> ANSWER (summarize):  

## Visualizing categorical vs categorical relationships {-}

Finally, suppose that instead of Trump's percentage support, we simply want to explore his county-level wins/losses:    

```{r echo=FALSE,fig.width=4, fig.height=4}
ggplot(elect, aes(x = winrep_2016)) +
  geom_bar()
```

Specifically, let's explore the relationship between `winrep_2016` and `StateColor`, another categorical variable.  

```{exercise name="Side-by-side bar plots"}
We saw above that we can incorporate a new categorical variable into a visualization by using grouping features such as color or facets.  Let's add information about `StateColor` to our bar plot of `winrep_2016`.    

```

a. Construct the following 4 bar plot visualizations.    

> ANSWER:  Run Code Below.

```{r eval=TRUE}
# a stacked bar plot
ggplot(elect, aes(x = StateColor, fill = winrep_2016)) +
  geom_bar()

# a side-by-side bar plot
ggplot(elect, aes(x = StateColor, fill = winrep_2016)) +
  geom_bar(position = "dodge")

# a proportional bar plot
ggplot(elect, aes(x = StateColor, fill = winrep_2016)) +
  geom_bar(position = "fill")

# faceted bar plot
ggplot(elect, aes(x = StateColor, fill = winrep_2016)) +
  geom_bar() +
  facet_wrap(~winrep_2016)
```    

b. Name one pro and one con of using the "proportional bar plot" instead of one of the other three options.    

> ANSWER:  

c. What's your favorite bar plot from part (a)?  Why?    

> ANSWER:  
   
## Practice {-}

### Hot Dogs {-}

In the annual Nathan's hot dog eating contest, people compete to eat as many hot dogs as possible in ten minutes.  Data on past competitions were compiled by Nathan Yau for "Visualize This: The FlowingData Guide to Design, Visualization, and Statistics": 

```{r, message=FALSE}
hotdogs <- read_csv("http://datasets.flowingdata.com/hot-dog-contest-winners.csv")
```   

```{exercise}
Address the following:
  
a. Construct a visualization of the winning number of hot dogs by year. THINK: Which is the response variable?      
b. Temporal trends are often visualized using a line plot.  Add a `geom_line()` layer to your plot from part (a).       
c. Summarize your observations about the temporal trends in the hot dog contest.    

```

> ANSWER (summary):  

```{r}
#a,b)


```



```{exercise}
All but two of the past winners are from the U.S. or Japan:

```

```{r}
table(hotdogs$Country)
```    

Use the following code to *filter* out just the winners from U.S. and Japan and name this `hotdogsSub`.  (Don't worry about the code itself - we'll discuss similar syntax later in the semester!)    

```{r}
library(dplyr)
hotdogsSub <- hotdogs %>%
  filter(Country %in% c("Japan", "United States"))
```    

a. Using a density plot approach *without* facets, construct a visualization of how the number of hot dogs eaten varies by country.
b. Repeat part a using a density plot approach *with* facets.    
c. Repeat part a using *something other than* a density plot approach.  (There are a few options!)    
d. Summarize your observations about the number of hot dogs eaten by country.    

> ANSWER (summarize):  

```{r}
#a)




#b)




#c)

```

### The Bechdel Test {-}

Recall the "Bechdel test" data from the previous activity.  As a reminder, the “Bechdel test” tests whether movies meet the following criteria:    

- there are $\ge$ 2 female characters    
- the female characters talk to each other    
- at least 1 time, they talk about something other than a male character    


```{r}
library(fivethirtyeight)
data(bechdel)
```
    
In investigating budgets and profits, the authors "focus on films released from 1990 to 2013, since the data has significantly more depth since then."  Use the following code to filter out just the movies in these years and name the resulting data set `Beyond1990` (don't worry about the syntax):    

```{r}
library(dplyr)
Beyond1990 <- bechdel %>%
  filter(year >= 1990)
```

```{exercise}
Address the following:
```

a. Construct a visualization that addresses the following research question: Do bigger budgets (`budget_2013`) pay off with greater box office returns (`domgross_2013`)?  In constructing this visualization, add a smooth to highlight trends and pay attention to which of these variables is the response.       
b. Using your visualization as supporting evidence, answer the research question.          
c. Part of the fivethirtyeight article focuses on how budgets (`budget_2013`) differ among movies with different degrees of female character development (`clean_test`).  Construct a visualization that highlights the relationship between these two variables.  There are many options - some are better than others!       
d. Using your visualization as supporting evidence, address fivethirtyeight's concerns.  

> ANSWER:  

```{r}
# a


# b


# c


# d
```

```{exercise}
NOTE: The following exercise is inspired by a similar exercise proposed by Albert Kim, one of the `fivethirtyeight` package authors.    
    Return to the fivethirtyeight.com article,["The Dollar-And-Cents Case Against Hollywood's Exclusion of Women"](http://fivethirtyeight.com/features/the-dollar-and-cents-case-against-hollywoods-exclusion-of-women/), and examine the plot titled "The Bechdel Test Over Time".    
```

a. Summarize the trends captured by this plot.  (How has the representation of women in movies evolved over time?)   
b. Recreate this plot from the article!  

To do so, you'll need to create a new data set named `newbechdel` in which the order of the Bechdel categories (`clean_test`) and the year categories (`yearCat`) match those used by fivethirtyeight.  Don't worry about the syntax: 

> ANSWER: a)


b) Add Code Below.

```{r}
library(dplyr)
newbechdel <- bechdel %>%
  mutate(clean_test = factor(bechdel$clean_test, c("nowomen", "notalk", "men", "dubious", "ok"))) %>%
  mutate(yearCat = cut(year, breaks = seq(1969, 2014, by = 5)))
```    
    
Further, you'll need to add the following layer in order to get a color scheme that's close to that in the article:   
    
```{r eval=FALSE}
scale_fill_manual(values = c("red", "salmon", "pink", "steelblue1", "steelblue4"))
```
    
```{r}
#b) 


```


 
