---
title: "Comp/Stat 112: Regular Expressions"
author: "INSERT STUDENT NAME HERE (COLLABORATOR NAMES HERE)"
output:
  bookdown::html_document2:
    split_by: none
    toc: yes
    toc_depth: 3
    toc_float:
      toc_collapsed: true
    number_sections: false
---

```{r 16_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(ggmap)
library(lubridate)
library(rvest)
library(stringr)
```

# Regular Expressions

## Learning Goals {-}

- Develop comfort in working with strings of text data 
- Use regular expressions to search and replace, detect patterns, locate patterns, extract patterns, and separate text with the `stringr` package.


## Regular Expressions and Character Strings {-}

Regular expressions allow us to describe character patterns. Regular expressions allow us to:^[Source: [regular expression tutorial](https://ryanstutorials.net/regular-expressions-tutorial/).]

- **Search** for particular items within a large body of text. For example, you may wish to identify and extract all email addresses.
- **Replace** particular items. For example, you may wish to clean up some poorly formatted HTML by replacing all uppercase tags with lowercase equivalents.
- **Validate** input. For example, you may want to check that a password meets certain criteria such as, a mix of uppercase and lowercase, digits and punctuation.
- **Coordinate** actions. For example, you may wish to process certain files in a directory, but only if they meet particular conditions.
- **Reformat** text. For example, you may want to split strings into different parts, each to form new variables.
- and more...

**Start by doing this [interactive tutorial](https://regexone.com/).** Note that neither the tutorial nor regular expressions more generally are specific to `R`. Some of the syntax in the tutorial is slightly different from what we'll use in `R`, but it will still help you get acclimated to the main ideas of regular expressions.

## Wrangling with Regular Expressions in `R` {-}

Now that we have some idea how regular expressions work, let's examine how to use them to achieve various tasks in `R`. It will be helpful to have your [cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/strings.pdf) handy. Many of these tasks can either be accomplished with functions from the `base` (built-in) package in `R` or from the `stringr` package, which is part of the Tidyverse. In general, the `stringr` functions are faster, which will be noticeable when processing a large amount of text data.

```{r}
example <- "The quick brown fox jumps over the lazy dog."
example2 <- "Two roads diverged in a yellow wood, / And sorry I could not travel both / And be one traveler, long I stood / And looked down one as far as I could"
example3 <- "This is a test"
```

### Search and replace patterns with `gsub` (`base`) or `str_replace` or `str_replace_all` (`stringr`) {-}

To search for a pattern and replace it, we can use the function `gsub`. Here are some examples:

```{r}
gsub("quick", "really quick", example)
gsub("fox|dog", "****", example)
gsub("(fox|dog).$", "****", example) # at end of sentence only, "." for any character
gsub("(fox|dog)\\.$", "****", example) # at end of sentence only, "\\." only for a period
gsub("the", "a", example) # case-sensitive only matches one
gsub("[Tt]he", "a", example) # will match either t or T; could also make "a" conditional on capitalization of t
```


In the `stringr` package, `str_replace` only replaces the first matched pattern, while `str_replace_all` replaces all. Here is a repetition of a few of the examples from above:

```{r}
str_replace(example, "[Tt]he", "a")
str_replace_all(example, "[Tt]he", "a")
str_replace_all(example, "fox|dog", "****")
```

### Detect patterns with `grep` and `grepl` (`base`) or `str_detect` (`stringr`) {-}

```{r}
examples <- c(example, example2, example3)
pat <- "[^aeiouAEIOU ]{3}" # Regular expression for three straight consonants. Note that I've excluded spaces as well
grepl(pat, examples) # Are there three consonants in a row?
grep(pat, examples) # Which entries contain this pattern?
str_detect(examples, pat) # same as grepl
```

### Locate patterns with `regexpr` or `str_locate` {-}

```{r}
regexpr(pat, example) # starting position and length of first match
str_locate(example, pat) # starting position and ending position of first match
```

Let's check the answer:
```{r}
substr(example, 23, 25)
```

### Extract patterns with `str_extract` and `str_extract_all` {-}

```{r}
pat2 <- "[^aeiouAEIOU ][aeiouAEIOU]{2}[^aeiouAEIOU ]{1}" # consonant followed by two vowels followed by a consonant
str_extract(example2, pat2) # extract first match
str_extract_all(example2, pat2, simplify = TRUE) # extract all matches
```

### Count the number of characters with `str_length` {-}

```{r}
str_length(example2)
```

### Convert a string to all lower case letters with `str_to_lower` {-}

```{r}
str_to_lower(example2)
```

### Split strings with `separate`  {-}

```{r}
df <- tibble(ex = example2)
df <- separate(df, ex, c("line1", "line2", "line3", "line4"), sep = " / ")
df$line1
df$line2
df$line3
df$line4
```


Note: The function `separate` is in the `tidyr` package.

## Practice: Fall 2021 Enrollment Exploration {-}

The tibble `courses` has the Fall 2021 enrollment information from the Macalester Registrar's website, which we scraped in an earlier activity.

```{r, echo=FALSE, cache=TRUE}
fall2021 <- read_html("https://www.macalester.edu/registrar/schedules/2021fall/class-schedule")

# Retrieve and inspect course numbers
course_nums <-
  fall2021 %>%
  html_nodes(".class-schedule-course-number") %>%
  html_text()

# Retrieve and inspect course names
course_names <-
  fall2021 %>%
  html_nodes(".class-schedule-course-title") %>%
  html_text()

course_nums_clean <- stringr::str_sub(course_nums, end = nchar(course_nums) - 6)

crn <- stringr::str_sub(course_nums, start = nchar(course_nums) - 4)

course_instructors <-
  fall2021 %>%
  html_nodes(".class-schedule-label:nth-child(6)") %>%
  html_text()
course_instructors_short <- stringr::str_sub(trimws(course_instructors), start = 13)

course_days <-
  fall2021 %>%
  html_nodes(".class-schedule-label:nth-child(3)") %>%
  html_text()
course_days_short <- stringr::str_sub(trimws(course_days), start = 7)

course_times <-
  fall2021 %>%
  html_nodes(".class-schedule-label:nth-child(4)") %>%
  html_text()
course_times_short <- stringr::str_sub(trimws(course_times), start = 7)

course_rooms <-
  fall2021 %>%
  html_nodes(".class-schedule-label:nth-child(5)") %>%
  html_text()
course_rooms_short <- stringr::str_sub(trimws(course_rooms), start = 7)

course_avail <-
  fall2021 %>%
  html_nodes(".class-schedule-label:nth-child(7)") %>%
  html_text()
course_avail_short <- stringr::str_sub(trimws(course_avail), start = 14)


CRNS <- courses %>% pull(crn)

SITES <- paste0("https://webapps.macalester.edu/registrardata/classdata/Fall2021/", CRNS) %>%
  purrr::map(~ read_html(.x))

course_desc <- SITES %>%
  purrr::map_chr(~ html_nodes(.x, "p:nth-child(1)") %>%
    html_text() %>%
    trimws())

gen_ed <- SITES %>%
  purrr::map_chr(~ html_nodes(.x, "p:nth-child(2)") %>%
    html_text() %>%
    trimws() %>%
    stringr::str_sub(start = 32) %>%
    trimws())


courses <-
  tibble(
    number = course_nums_clean,
    crn = crn,
    name = course_names,
    days = course_days_short,
    time = course_times_short,
    room = course_rooms_short,
    instructor = course_instructors_short,
    avail_max = course_avail_short,
    desc = course_desc,
    gen_ed = gen_ed
  )
```

```{r, echo=FALSE}
knitr::kable(
  courses[1:6, 1:8],
  caption = "First six entries in the Fall 2021 Macalester course info data."
)
```





```{exercise,name="Rearrange data table"}
Make the following changes to the `courses` data table:
```  


a) Split `number` into three separate columns: `dept`, `number`, and `section`.

> ANSWER:

```{r}

```

b) Split the `avail_max` variable into two separate variables: `avail` and `max`. It might be helpful to first remove all appearances of "Closed ". 

> ANSWER:

```{r}

```

c) Use `avail` and `max` to generate a new variable called `enrollment`.

> ANSWER:

```{r}

```

d) Split the `time` variable into two separate columns: `start_time` and `end_time`. Convert all of these times into continuous 24 hour times (e.g., 2:15 pm should become 14.25). *Hint: check out the function `parse_date_time`.*

> ANSWER:

```{r}

```






```{exercise,name="WA courses"}
Make a histogram showing the number of Fall 2021 sections satisfying the Writing WA requirement, sorted by department code.^[For this exercise, you can count cross-listed courses towards both departments' WA counts.]

```

> ANSWER:

```{r}

```

<br>

In the next series of exercises, we are going to build up an analysis to examine the number of student enrollments for each faculty member.

```{exercise, name="Filter cases"}
For this particular analysis, we do not want to consider certain types of sections. Remove all of the following from the data table:

a) All sections in `PE` or `INTD`.
b) All music ensembles and dance practicum sections (these are all of the MUSI and THDA classes with numbers less than 100).
c) All lab sections. This is one is a bit tricky. You can search for "Lab" or "Laboratory", but be careful not to eliminate courses with words such as "Labor". Some of these have section numbers that end in "-L1"", for example.

```

> ANSWER:

```{r}

```


```{exercise, label="nt",name="Handle cross-listed courses"}
Some sections are listed under multiple different departments, and you will find the same instructor, time, enrollment data, etc. For this activity, we only want to include each actual section once and it doesn't really matter which department code we associate with this section. Eliminate all duplicates, keeping each actual section just once. Hint: look into the `R` command `distinct`, and think carefully about how to find duplicates.

```

> ANSWER:

```{r}

```


```{exercise, name="Co-taught courses"}
Make a table with all Fall 2021 co-taught courses (i.e., more than one instructor).

```

> ANSWER:

```{r}

```


```{exercise, name="Faculty enrollments"}
Make a table where each row contains a faculty, the number of sections they are teaching in Fall 2021, and the total enrollments in those section. Sort the table from highest total enrollments to lowest.^[For the purposes of this exercise, we are just going to leave co-taught courses as is so that you will have an extra row for each pair or triplet of instructors. Alternatives would be to allocate the enrollment number to each of the faculty members or to split it up between the members. The first option would usually be the most appropriate, although it might depend on the course.]

```

> ANSWER:

```{r}

```


```{exercise, name="Evening courses"}
Create and display a new table with all night courses (i.e., a subset of the table you wrangled by the end of Exercise \@ref(exr:nt)). Also make a histogram showing the number of these courses by day of the week.

```

> ANSWER:

```{r}

```
