```{r 10_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, tidy = FALSE, message = FALSE, cache.extra = packageVersion("tufte"))
library(tidyverse)
library(ggplot2)
```

# Categorical Variables as Factors

## Learning Goals {-}

- Understand the difference between a variable stored as a `character` vs. a `factor`
- Be able to convert a `character` variable to a `factor`
- Develop comfort in manipulating the order and values of a factor with the `forcats` package to improve summaries and visualizations.

You can download a template .Rmd of this activity [here](template_rmd/10-Factors_Assign.Rmd). Put this in a folder `Day_10` in your `COMP_STAT_112` folder.

## Example: Grade Distribution {-}

```{r,cache=TRUE,echo=FALSE}
Grades <- read_csv("data/grades.csv")
Grades <- Grades %>%
  select(sid, sessionID, grade) %>%
  distinct(sid, sessionID, .keep_all = TRUE)
```

```{r,cache=TRUE,echo=TRUE,eval=FALSE}
Grades <- read_csv("https://bcheggeseth.github.io/112_fall_2022/data/grades.csv")
Grades <- Grades %>%
  select(sid, sessionID, grade) %>%
  distinct(sid, sessionID, .keep_all = TRUE)
```

We will continue with the grades data from the previous activity. Here is a sample to remember what it looks like:

```{r, echo=FALSE}
set.seed(101)
knitr::kable(
  sample_n(Grades, size = 4),
  align = "ccc", caption = "Student grades."
)
```

Here is a bar chart of the grade distribution:

```{r}
ggplot(Grades, aes(x = grade)) +
  geom_bar()
```

We can also wrangle a table that just has each grade and the number of times it appears:

```{r}
GradeDistribution <- Grades %>%
  group_by(grade) %>%
  summarize(count = n())
```

```{r eval=FALSE}
# Alternatively, we can use the count() function the creates a variable called n
Grades %>%
  count(grade) 
```

```{r, echo=FALSE}
knitr::kable(
  GradeDistribution,
  caption = "Grade distribution."
)
```


***What could be improved about this graphic and table?***

The grades are listed alphabetically, which isn't particularly meaningful. Why are they listed that way? Because the variable `grade` is a character string type:

```{r}
class(Grades$grade)
```

When dealing with categorical variables that take a finite number of values (`levels`, formally), it is often useful to store the variable as a factor, and specify a meaningful order for the levels. 

For example, when the entries are stored as character strings, we cannot use the `levels` command to see the full list of values:

```{r}
levels(Grades$grade)
```

### Converting to factor {-}

Let's first convert the `grade` variable to a factor:

```{r}

Grades <- Grades %>%
  mutate(grade = factor(grade))
```

Now we can see the levels:

```{r}
levels(Grades$grade)
```

Moreover, the `forcats` package (part of `tidyverse`) allows us to manipulate these factors. Its commands include the following. 

### Changing the order of levels {-}
   
- `fct_relevel()`: manually reorder levels   
- `fct_infreq()`: order levels from highest to lowest frequency   
- `fct_reorder()`: reorder levels by values of another variable   
- `fct_rev()`: reverse the current order   

### Changing the value of levels {-} 

- `fct_recode()`: manually change levels   
- `fct_lump()`: group together least common levels   

More details on these and other commands can be found on the [forcats cheat sheet](https://github.com/rstudio/cheatsheets/raw/main/factors.pdf) or in Wickham & Grolemund's [chapter on factors](https://r4ds.had.co.nz/factors.html).

\

```{example, name="Reorder factors"}
Let's reorder the grades so that they are in a more meaningful order for the bar chart above. Here are three options:
   
```

\

**Option 1**: From high grade to low grade, with "S" and "AU" at the end:

```{r}
Grades %>%
  mutate(grade = fct_relevel(grade, c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "NC", "S", "AU"))) %>%
  ggplot(aes(x = grade)) +
  geom_bar()
```

**Option 2**: In terms of ascending frequency:

```{r}
ggplot(GradeDistribution) +
  geom_col(aes(x = fct_reorder(grade, count), y = count)) +
  labs(x = "grade")
```
     
**Option 3**: In terms of descending frequency:

```{r}
ggplot(GradeDistribution) +
  geom_col(aes(x = fct_reorder(grade, count, .desc = TRUE), y = count)) +
  labs(x = "grade")
```        
         

```{example, name="Recode factors"}
Because it may not be clear what "AU" and "S" stand for, let's rename them to "Audit" and "Satisfactory".   
```


```{r}
Grades %>%
  mutate(grade = fct_relevel(grade, c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "NC", "S", "AU"))) %>%
  mutate(grade = fct_recode(grade, "Satisfactory" = "S", "Audit" = "AU")) %>%
  ggplot(aes(x = grade)) +
  geom_bar()
```

```{exercise}
Now that you've developed your data visualization and wrangling skills, 
``` 

a. develop a research question to address with the grades and courses data,
b. create a high quality visualization that addresses your research question,
c. write a brief description of the visualization and include the insight you gain about the research question.

```
Courses <- read_csv("https://bcheggeseth.github.io/112_fall_2022/data/courses.csv")
```

## Appendix: R Functions {.unnumbered}

### Changing the order of levels {.unnumbered}

| Function/Operator |                 Action                 |                       Example |
|-------------------------|:---------------------------:|------------------:|
| `fct_relevel()`        |  manually reorder levels of a factor  |   `Grades %>% mutate(grade = fct_relevel(grade, c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "NC", "S", "AU")))` |
| `fct_infreq()`        | order levels from highest to lowest frequency    |   `ggplot(Grades) + geom_bar(aes(x = fct_infreq(grade)))` |
| `fct_reorder()`        | reorder levels by values of another variable |   `ggplot(GradeDistribution) + geom_col(aes(x = fct_reorder(grade, count), y = count))` |
| `fct_rev()`        | reverse the current order |   `ggplot(Grades) + geom_bar(aes(x = fct_rev(fct_infreq(grade))))` |



### Changing the value of levels {-} 

| Function/Operator |                 Action                 |                       Example |
|-------------------------|:---------------------------:|------------------:|
| `fct_recode()`        |  manually change levels    |   `Grades %>%  mutate(grade = fct_recode(grade, "Satisfactory" = "S", "Audit" = "AU")) ` |
| `fct_lump()`        | group together least common levels      |   `Grades %>% mutate(grade = fct_lump(grade, n = 5))` |

